---
title: "Assignment 3: Diamonds Price Estimation"
author: "ilaydacelenk_IE48A"
date: "8/25/2020"
output: 
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    
---

## Introduction
This analysis focuses on the relationship between price and quality information of diamonds. The main objective is to create a model which predicts the price of a given diamond. 
The `diamonds` dataset is included in the `ggplot2` package. Therefore it is necessary to load it. 

## Loading Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#set.seed(503)
library(dplyr)
library(ggplot2)
```

## Overview and Manipulation of Data
Let's take a quick view of the dataset. There are `r diamonds %>% summarise(count=n())` observations and `r length(diamonds)` variables.

```{r dataset}
str(diamonds)
```

### Variables

* __carat__ : weight of the diamond between (`r min(diamonds$carat)` ,`r max(diamonds$carat)`)
* __cut__ : quality of the cut in (`r levels(diamonds$cut)`) from worst to best
* __color__ : diamond color in (`r levels(diamonds$color)`) from worst to best
* __clarity__ : clarity of the diamond in (`r levels(diamonds$clarity)`) from worst to best
* __depth__ : total depth percentage = z / mean(x, y) between (`r min(diamonds$depth)` ,`r max(diamonds$depth)`).
* __table__ : width of top of diamond relative to widest point between (`r min(diamonds$table)` ,`r max(diamonds$table)`).
* __price__ : price in US dollars between (`r min(diamonds$price)` ,`r max(diamonds$price)`)
* __x__ : length in mm between (`r min(diamonds$x)` ,`r max(diamonds$x)`)
* __y__ : width in mm between (`r min(diamonds$y)` ,`r max(diamonds$y)`)
* __z__ : depth in mm between (`r min(diamonds$z)` ,`r max(diamonds$z)`).

The variables `cut`, `color` and `clarity` are ordinal categorical variables, which means they are categorical but they can be ordered as factors. Therefore their levels are written from worst to best.
<br> Figure below gives some information on `carat`, `cut`, `color` and `clarity`. 

```{r }

url <- "https://raw.githubusercontent.com/pjournal/boun01-ilaydacelenk/gh-pages/Assignment3/cut-color-clarity-carat.jpeg"
z <- tempfile()
download.file(url,z,mode="wb")
img <- readJPEG(z)
file.remove(z)
img
```


```{r }

diamonds_test <- diamonds %>% mutate(diamond_id = row_number()) %>% 
    group_by(cut, color, clarity) %>% sample_frac(0.2) %>% ungroup()

diamonds_train <- anti_join(diamonds %>% mutate(diamond_id = row_number()), 
    diamonds_test, by = "diamond_id")


diamonds_train
diamonds_test
```








## References
- [Assignment page](https://mef-bda503.github.io/archive/fall17/files/assignment_diamonds_data.html)


<br>To see all my work related to IE 48A course, visit my [Progress Journal website](https://pjournal.github.io/boun01-ilaydacelenk/). You can also visit my personal Github [Website](https://ilaydacelenk.github.io/).